Normalization by Evaluation for Sized Dependent Types

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Missing resurrects and missing principal types.

\item We present the first integration of a dependent type theory with higher-rank size polymorphism.
Concretely, we consider a type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).  Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
(In the following rules, $a$, $a'$, and $b$ stand for arbitrary size expressions.)
\[
\ru{\Gamma \der t = t' : \forall i.\, T
  }{\Gamma \der t\,a = t\,a' : T[b/i]}
\qquad
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[b/i]}
\qquad
\qquad
\ru{\Gamma \der t = t' : \Pi i.\, T
  }{\Gamma \der t\,a = t\,a : T[a/i]}
\qquad
\ru{\Gamma \der t : \Pi i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]

@@

\noindent
In the following, we list syntactic properties of our judgements.  To
this end, let $J$ %, $L$, and $R$
match a part of a judgement.

@@

%A type constructor $B \in \FixK\,\ell$ is \emph{admissible for recursion}, \fbox{$B \in \ADM$}, if
%$\bigcap_{
\noindent
Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
$B\,\beta\, a \leq B\,\infty\,a \in \TYPE$ and
$B'\,\beta\, a \leq B'\,\infty\,a \in \TYPE$.
\begin{lemma}[Fix]
Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
If\/ $a = a' \in \Nat \alpha$
and $B = B' \in \Adm\ell$
and $f = f' \in \FixT\,B$
then $g = g' \in B\,\alpha\,a$.
\end{lemma}
\begin{proof*}
By well-founded induction on $\alpha$.
\begin{caselist}

@@

In this section we show that the declarative judgements are sound, in
particular, well-formed syntactic types map to semantic types, and
definitionally equal terms map to related values in the PER model.
The proof is standard.  First, we define inductively a
PER of substitutions
\fbox{$\eta = \eta' \eeq \rho \in \Gamma$}.
\begin{gather*}
  \ru{
    }{\sempty = \sempty \eeq \sempty \in \cempty}
\qquad
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad
      T\rho \in \SET[\ell] \qquad
      u = u' = t \in T \rho
    }{\sext \eta u = \sext{\eta'}{u'} \eeq \sext{\rho}{t} \in \cext \Gamma T}
\\[2ex]
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha \in \SIZE
    }{\sext \eta \alpha = \sext{\eta'}{\alpha} \eeq \sext{\rho}{\alpha} \in \cext \Gamma \Size}
\qquad
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha,\alpha',\beta \in \SIZE
    }{\sext \eta \alpha = \sext{\eta'}{\alpha'} \eeq \sext{\rho}{\beta} \in \erext \Gamma \Size}
\end{gather*}
We write $\rho \in \Gamma$ for $\rho = \rho \eeq \rho \in \Gamma$.
\begin{lemma}[Resurrection]
  \label{lem:resenv}
  If\/ $\eta = \eta' \eeq \rho \in \Gamma$ then $\rho \in \resurrect\Gamma$.
\end{lemma}

Then, we define semantic counterparts of our declarative judgements:
\[
\begin{array}{l@{\quad}l@{\quad}l}
  \models () & \defas & \mtrue \\
  \models \eext \Gamma \Size & \defas & \models \Gamma \\
  \models \cext \Gamma T     & \defas & \models \Gamma \mand \resurrect\Gamma \models T
\\[2ex]
  \Gamma \models s & \defas & \models \Gamma \\
%  \Gamma \models s & \defas & \mtrue \\
  \Gamma \models T & \defas & \Gamma \models T = T \\
  \Gamma \models T = T' & \defas & \Gamma \models T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
\\[2ex]
  \Gamma \models T \leq T' & \defas & \Gamma \models T \mand \Gamma \models T' \mand
    T\eta \leq T'\eta' \in \TYPE
    % \EL(T\eta) \subseteq \EL(T'\eta')
    \mforall \eta = \eta' \eeq \rho \in \Gamma
\\[2ex]
  \Gamma \models t : T & \defas & \Gamma \models t = t : T \\
  \Gamma \models t = t' : T & \defas & %\models \Gamma \mand
    % \resurrect\Gamma \models T \mand
    \models \cext \Gamma T \mand
    t \eta = t' \eta' \in T \rho \mforall \eta = \eta' \eeq \rho \in \Gamma
\\[2ex]
  \Gamma \models \sigma : \Delta & \defas & \Gamma \models \sigma = \sigma \eeq \sigma : \Delta \\
  \Gamma \models \sigma = \sigma' \eeq \tau : \Delta & \defas &
    \models \Gamma \mand \models \Delta \mand
    \sigma \eta = \sigma' \eta' \eeq \tau \rho \in \Delta
    \mforall \eta = \eta' \eeq \rho \in \Gamma
\end{array}
\]

\begin{theorem}[Fundamental theorem]
  \label{thm:fund}
  \bla
  \begin{enumerate}
  \item
  If\/ $\der \Gamma$ then $\models \Gamma$.
  \item
  If\/ $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$
  \end{enumerate}
\end{theorem}
\SHORTVERSION{
\begin{proof}
Simultaneously, by induction on the derivation.
\end{proof}
}
\LONGVERSION{  % BEGIN LONGVERSION
\begin{proof*}
Simultaneously, by induction on the derivation.
% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).

@@

\nextcase
\vspace{-3ex}
\[
  \ru{\der \Gamma
    }{\der \eext\Gamma \Size}
\]
By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size)$.
\qed
\end{caselist}
\end{proof*}

@@

  \noindent
  $\LRTS{\Gamma}{T'}{A'} s$ iff
  $\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
  $\Gamma \der T' \doteq \tRty\,A' : s$
  %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$
  % $\forall \xi : \Gamma' \leq \Gamma.\ \
  %  \exists m. \ \
  %  \R {\Gamma'} {\down {A'} f} m \mand
  %  \Gamma' \der t\xi = m : T'\xi$.

@@

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' \evalsto \Nat a$ for some $a$ and
  $\LRSize {\resurrect\Gamma} a \alpha$ and
  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$
  % \\
  % $\forall \xi : \Gamma' \leq \Gamma.\ \
  %  \exists v. \ \
  %  \R {\Gamma'} {\down {A'} f} v \mand
  %  \Gamma' \der t\xi = v : \Nat\,a\xi$.

@@

\noindent
$\LRTS \Gamma {T'} {A'} s$ iff
\(
\begin{array}[t]{l}
\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma,\
{\Gamma'} \der b : \Size,\
\beta \in \SIZE. \ \
\LRSize {\Gamma'} b \beta
\implies \LRTS {\Gamma'} {T(\xi,b)} {B[\beta]} s
.
\end{array}
\) \\
$\LR \Gamma t {T'} f {A'}$ iff
\(
\begin{array}[t]{l}
\resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma,\
%\Gamma' \der a : \Size,\
\resurrect{\Gamma'} \der a,b : \Size,\
\alpha,\beta \in \SIZE. \ \\ \qquad
\LRSize {\resurrect{\Gamma'}} b \beta
\implies \LR {\Gamma'} {t\xi\,\ann a} {T(\xi,b)} {f\,\ann\alpha} {B[\beta]}
.
\end{array}
\)
\end{caselist}
We may prove theorems ``by induction on $\LRTS \Gamma T A s$'', even if in reality this will be proofs by induction on $A \in s$ and cases on $\LRTS \Gamma T A s$.
We write \fbox{$\RG \Gamma T A$} if\/ $\LRTS \Gamma T A s$ for some sort $s$.
\SHORTVERSION{The logical relations are closed under weakening.}
\LONGVERSION{ % BEGIN LONG
\begin{lemma}[Weakening]
  Let $\xi : \Gamma' \leq \Gamma$.
  \begin{enumerate}
  \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
  \item If\/ $\LRTS {\Gamma} T A s$ then $\LRTS{\Gamma'}{T\xi}A s$.
  \item If\/ $\LR \Gamma t T f A$ then $\LR{\Gamma'}{t\xi}{T\xi} f A$.
  \end{enumerate}
\end{lemma}
} % END LONG

@@

\begin{theorem}[Fundamental theorem of typing]
\label{thm:fundty}
\bla
\begin{enumerate}
\item
If\/ $\Gamma \der t : T$ then $\Gamma \valid t : T$.
\item
If\/ $\Gamma \der \sigma : \Delta$ then
$\Gamma \valid \sigma : \Delta$
\end{enumerate}
\end{theorem}
\begin{proof*}
Each by induction on the derivation.
\SHORTLONG{\qed}{ % LONGVERSION
\begin{caselist}

@@

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t : \forall T \qquad % \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der a,b : \Size
    }{\Gamma \der t \ann a : T[b]}
\]
Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
$\LR \Delta {t \ann a \sigma} {T[b]\tau} {t \ann a \eta} {T[b]\rho}$.
By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
It suffices to show
\begin{enumerate}
\item $\resurrect\Delta \der a\sigma : \Size$: follows from $\resurrect\Gamma \der a : \Size$
  via $\Delta \der \sigma : \Gamma$ and $\resurrect\Delta \der \sigma : \resurrect\Gamma$
  and substitution.
%\item $\resurrect\Delta \der b\tau : \Size$.
\item $a\eta \in \SIZE$: follows from $\resurrect\Gamma \der a : \Size$
  by Lemma~\ref{lem:wklrsub}, part (\ref{it:sizesub}).
%\item $b\rho \in \SIZE$.
\item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
\qed
\end{enumerate}
\end{caselist}
} % END LONGVERSION
\end{proof*}

@@

% Reification erases all irrelevant size arguments to $\infty$.  We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.  However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.  Proving this statement seems to be the key to an incremental equality check; but
% the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.

@@

Finally, we would like a general theory of shape-irrelevance that extends beyond size-indexed types.  For instance, any data type constructor could be considered shape-irrelevant in all its parameters.  However, the current reification algorithm which erases irrelevant size arguments to $\infty$ would need some patching.  \citet{abelCoquandPagano:lmcs11} erase all proofs of a proposition $P$ to a canonical element $\star$ which inhabits any proposition $P$ that is already inhabited.  A similar technique could work in our setting and would be necessary unless we can justify the incremental equality check (see above).

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  Swedish Research Council (Vetenskapsr\aa{}det)
  under Grant
  No.~621-2014-4864 \emph{Termination Certificates for Dependently-Typed Programs and Proofs via Refinement Types}.
  The first author is grateful for recent discussions with Thierry Coquand, Nils Anders Danielsson, and Sandro Stucki
  which helped clarifying the thoughts leading to this work.
  He also acknowledges past discussions with Christoph-Simon Senjak.
\end{acks}

>>>>

\item We present the first integration of a dependent type theory with higher-rank size polymorphism.
Concretely, we consider a type theory \`a la Martin-L\"of with dependent function types, cumulative universes, subtyping, a judgmental equality with $\eta$-laws, a sized type of natural numbers and two size quantifiers: an irrelevant one $(\forall)$ for binding of sizes in irrelevant positions, and a relevant one $(\Pi)$ for binding of sizes in shape-irrelevant positions (Section~\ref{sec:syntax}).  Judgmental equality features a ``Curry-style'' rule for irrelevant size application which ignores the size arguments, and consequently, the corresponding typing rule will also ignore the size argument.
(In the following rules, $a$, $a'$, and $b$ stand for arbitrary size expressions.)
\[
\ru{\Gamma \der t = t' : \forall i.\, T
  }{\Gamma \der t\,a = t\,a' : T[b/i]}
\qquad
\ru{\Gamma \der t : \forall i.\, T
  }{\Gamma \der t\,a : T[b/i]}
\qquad
\qquad
\ru{\Gamma \der t = t' : \Pi i.\, T
  }{\Gamma \der t\,a = t\,a : T[a/i]}
\qquad
\ru{\Gamma \der t : \Pi i.\, T
  }{\Gamma \der t\,a : T[a/i]}
\]
Our substitution theorem distinguishes term- from type-side substitutions.

@@

\noindent
We should remark here that we have \emph{neither type unicity nor principal types}
due to the irrelevant size application rule.
In the following, we list syntactic properties of our judgements.  To
this end, let $J$ %, $L$, and $R$
match a part of a judgement.

@@

%A type constructor $B \in \FixK\,\ell$ is \emph{admissible for recursion}, \fbox{$B \in \ADM$}, if
%$\bigcap_{
\noindent
As the semantic counterpart of judgement $\Gamma \der T : \Adm \ell$,
let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
$B\,\beta\, a \leq B\,\infty\,a \in \TYPE$ and
$B'\,\beta\, a \leq B'\,\infty\,a \in \TYPE$.
\begin{lemma}[Fix]
Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
If\/ $a = a' \in \Nat \alpha$
and $B = B' \in \Adm\ell$
and $f = f' \in \FixT\,B$
then $g = g' \in B\,\alpha\,a$.
\end{lemma}
\begin{proof*}
By well-founded induction on $\alpha$.
\begin{caselist}

@@

In this section we show that the declarative judgements are sound, in
particular, well-formed syntactic types map to semantic types, and
definitionally equal terms map to related values in the PER model.
The proof runs the usual course.  First, we define inductively a
PER of substitutions
\fbox{$\eta = \eta' \eeq \rho \in \Gamma$}\,.
\begin{gather*}
  \ru{
    }{\sempty = \sempty \eeq \sempty \in \cempty}
\qquad
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad
      T\rho \in \SET[\ell] \qquad
      u = u' = t \in T \rho
    }{\sext \eta u = \sext{\eta'}{u'} \eeq \sext{\rho}{t} \in \cext \Gamma T}
\\[2ex]
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha \in \SIZE
    }{\sext \eta \alpha = \sext{\eta'}{\alpha} \eeq \sext{\rho}{\alpha} \in \cext \Gamma \Size}
\qquad
  \ru{\eta = \eta' \eeq \rho \in \Gamma \qquad \alpha,\alpha',\beta \in \SIZE
    }{\sext \eta \alpha = \sext{\eta'}{\alpha'} \eeq \sext{\rho}{\beta} \in \erext \Gamma \Size}
\end{gather*}
We write $\rho \in \Gamma$ for $\rho = \rho \eeq \rho \in \Gamma$.
\begin{lemma}[Resurrection]
  \label{lem:resenv}
  If\/ $\eta = \eta' \eeq \rho \in \Gamma$ then $\rho \in \resurrect\Gamma$.
\end{lemma}

\noindent
Then, we define semantic counterparts of our declarative judgements:
\[
\begin{array}{l@{\quad}l@{\quad}l}
  \models () & \defas & \mtrue \\
  \models \eext \Gamma \Size & \defas & \models \Gamma \\
  \models \cext \Gamma s     & \defas & \models \Gamma \\
  \models \cext \Gamma T     & \defas & \models \Gamma \mand \resurrect\Gamma \models T
\\[2ex]
  \Gamma \models s & \defas & \models \Gamma \\
%  \Gamma \models s & \defas & \mtrue \\
  \Gamma \models T & \defas & \Gamma \models T = T \\
  \Gamma \models T = T' & \defas & \Gamma \models T = T' : s \mforsome s %\Set[\ell] \mforsome \ell
\\[2ex]
  \Gamma \models T : \Adm\,\ell & \defas & \Gamma \models T : \FixK\,\ell \mand
    T\eta = T'\eta' \in \Adm\,\ell
    % \EL(T\eta) \subseteq \EL(T'\eta')
    \mforall \eta = \eta' \eeq \rho \in \Gamma
\\
  \Gamma \models T \leq T' & \defas & \Gamma \models T \mand \Gamma \models T' \mand
    T\eta \leq T'\eta' \in \TYPE
    % \EL(T\eta) \subseteq \EL(T'\eta')
    \mforall \eta = \eta' \eeq \rho \in \Gamma
\\[2ex]
  \Gamma \models t : T & \defas & \Gamma \models t = t : T \\
  \Gamma \models t = t' : T & \defas & %\models \Gamma \mand
    % \resurrect\Gamma \models T \mand
    \models \cext \Gamma T \mand
    t \eta = t' \eta' \in T \rho \mforall \eta = \eta' \eeq \rho \in \Gamma
\\[2ex]
  \Gamma \models \sigma : \Delta & \defas & \Gamma \models \sigma = \sigma \eeq \sigma : \Delta \\
  \Gamma \models \sigma = \sigma' \eeq \tau : \Delta & \defas &
    \models \Gamma \mand \models \Delta \mand
    \sigma \eta = \sigma' \eta' \eeq \tau \rho \in \Delta
    \mforall \eta = \eta' \eeq \rho \in \Gamma
\end{array}
\]

\begin{theorem}[Fundamental theorem]
  \label{thm:fund}
  \bla
  \begin{enumerate}
  \item
  If\/ $\der \Gamma$ then $\models \Gamma$.
  \item
  If\/ $\Gamma \der J$ then % $\models \Gamma$ and
  $\Gamma \models J$.
  \end{enumerate}
\end{theorem}
\SHORTVERSION{
\begin{proof}
Simultaneously, by induction on the derivation.
\end{proof}
}
\LONGVERSION{  % BEGIN LONGVERSION
\begin{proof*}
Simultaneously, by induction on the derivation.
% The soundness of subtyping and the conversion rule is due to subsumption (lemmata \ref{lem:subsump} and \ref{lem:usubsump}).

@@

\nextcase
\vspace{-3ex}
\[
  \ru{\der \Gamma
    }{\der \eext\Gamma \Size}
\]
By induction hypothesis $\rho_\Gamma \in \Gamma$.  Since $\var\Gamma \in \SIZE$ we conclude
$(\rho_\Gamma, \var\Gamma) \in \eext\Gamma \Size$.
\qed
\end{caselist}
\end{proof*}

@@

  \noindent
  $\LRTS{\Gamma}{T'}{A'} s$ iff
  $\Gamma \der T' \evalsto n : s$ for some neutral $n$ and
  $\Gamma \der T' \doteq \tRty\,A' : s$.
  %$\forall \xi : \Gamma' \leq \Gamma.\ \ \exists M.\ \ \RTy{\Gamma'}{A'}M \mand \Gamma' \der T'\xi = M : \Set[\ell]$.

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff
  $\Gamma \der t \doteq \tR \, \down {A'} f : T'$.
  % $\forall \xi : \Gamma' \leq \Gamma.\ \
  %  \exists m. \ \
  %  \R {\Gamma'} {\down {A'} f} m \mand
  %  \Gamma' \der t\xi = m : T'\xi$.

@@

  \noindent
  $\LR{\Gamma}{t}{T'}{f}{A'}$ iff $\resurrect\Gamma \der T' \evalsto \Nat a$ for some $a$ and
  $\LRSize {\resurrect\Gamma} a \alpha$ and
  $\Gamma \der t \doteq \tR\,\down{A'} f : \Nat a$.
  % \\
  % $\forall \xi : \Gamma' \leq \Gamma.\ \
  %  \exists v. \ \
  %  \R {\Gamma'} {\down {A'} f} v \mand
  %  \Gamma' \der t\xi = v : \Nat\,a\xi$.

@@

\noindent
$\LRTS \Gamma {T'} {A'} s$ iff
\(
\begin{array}[t]{l}
\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma,\
{\Gamma'} \der b : \Size,\
\beta \in \SIZE. \ \
\LRSize {\Gamma'} b \beta
\implies \LRTS {\Gamma'} {T(\xi,b)} {B[\beta]} s
.
\end{array}
\) \\
$\LR \Gamma t {T'} f {A'}$ iff
\(
\begin{array}[t]{l}
\resurrect\Gamma \der T' \evalsto \forallT T : s \mforsome T\\
\forall \xi : \Gamma' \leq \Gamma,\
%\Gamma' \der a : \Size,\
\resurrect{\Gamma'} \der a,b : \Size,\
\alpha,\beta \in \SIZE. \ \\ \qquad
\LRSize {\resurrect{\Gamma'}} b \beta
\implies \LR {\Gamma'} {t\xi\,\ann a} {T(\xi,b)} {f\,\ann\alpha} {B[\beta]}
.
\end{array}
\)
\end{caselist}
We may prove theorems ``by induction on $\LRTS \Gamma T A s$'', even if in reality this will be proofs by induction on $A \in s$ and cases on $\LRTS \Gamma T A s$.
We write \fbox{$\RG \Gamma T A$} if\/ $\LRTS \Gamma T A s$ for some sort $s$.
\SHORTVERSION{The logical relations are closed under weakening.}
\LONGVERSION{ % BEGIN LONG
\begin{lemma}[Weakening]
  Let $\xi : \Gamma' \leq \Gamma$.
  \begin{enumerate}
  %% SEE ABOVE
  % \item If\/ $\LRSize \Gamma a \alpha$ then $\LRSize{\Gamma'}{a\xi}\alpha$.
  \item If\/ $\LRTS {\Gamma} T A s$ then $\LRTS{\Gamma'}{T\xi}A s$.
  \item If\/ $\LR \Gamma t T f A$ then $\LR{\Gamma'}{t\xi}{T\xi} f A$.
  \end{enumerate}
\end{lemma}
} % END LONG

@@

\begin{theorem}[Fundamental theorem of typing]
\label{thm:fundty}
\bla
\begin{enumerate}
\item
If\/ $\Gamma \der t : T$ then $\Gamma \valid t : T$.
\item
If\/ $\Gamma \der \sigma : \Delta$ then
$\Gamma \valid \sigma : \Delta$.
\end{enumerate}
\end{theorem}
\begin{proof*}
Each by induction on the derivation.
\SHORTLONG{\qed}{ % LONGVERSION
\begin{caselist}

@@

\nextcase $\forall$-elimination.
\[
  \ru{\Gamma \der t : \forall T \qquad % \Gamma \der a : \Size \qquad
      \resurrect \Gamma \der a,b : \Size
    }{\Gamma \der t \ann a : T[b]}
\]
Assume $\LRSub {\Delta} {\sigma} {\tau} \Gamma {\eta} {\rho}$ and show
$\LR \Delta {t \ann a \sigma} {T[b]\tau} {t \ann a \eta} {T[b]\rho}$.
By induction hypothesis we have $\LR \Delta {t \sigma} {(\forall T)\tau} {t \eta} {(\forall T) \rho}$.
It suffices to show the following:
\begin{enumerate}
\item $\resurrect\Delta \der a\sigma : \Size$: follows from $\resurrect\Gamma \der a : \Size$
  via $\Delta \der \sigma : \Gamma$ and $\resurrect\Delta \der \sigma : \resurrect\Gamma$
  and substitution.
%\item $\resurrect\Delta \der b\tau : \Size$.
\item $a\eta \in \SIZE$: follows from $\resurrect\Gamma \der a : \Size$
  by Lemma~\ref{lem:wklrsub}, part (\ref{it:sizesub}).
%\item $b\rho \in \SIZE$.
\item $\LRSize {\resurrect \Delta} {b\tau} {b\rho}$: follows from induction hypothesis on $\resurrect \Gamma \der b : \Size$ via $\LRSub {\resurrect \Delta} \tau \tau {\resurrect\Gamma} \rho \rho$ (from Lemma~\ref{lem:wklrsub}).
\qed
\end{enumerate}
\end{caselist}
} % END LONGVERSION
\end{proof*}

@@

% Reification erases all irrelevant size arguments to $\infty$.  We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.  However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.  Proving this statement seems to be the key to an incremental equality check; but
% the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types, and even principal types.

@@

Finally, we would like a general theory of shape-irrelevance that
extends beyond size-indexed types.  For instance, any data type
constructor could be considered shape-irrelevant in all its
indices, with the consequence that index arguments in the data constructors
could be declared irrelevant.
However, our notion of judgmental equality does not support
irrelevant arguments of dependent type.
It works for the non-dependent type $\Size$, but we also
relied on having a closed inhabitant $\infty$ in $\Size$.
More research is needed to tell a more general story of shape-irrelevance.

% Finally, we would like a general theory of shape-irrelevance that
% extends beyond size-indexed types.  For instance, any data type
% constructor could be considered shape-irrelevant in all its
% parameters.
% However, the current reification algorithm which erases
% irrelevant size arguments to $\infty$ would need some patching.
% \citet{abelCoquandPagano:lmcs11} erase all proofs of a proposition $P$
% to a canonical element $\star$ which inhabits any proposition $P$ that
% is already inhabited.  A similar technique could work in our setting
% and would be necessary unless we can justify the incremental equality
% check (see above).

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  Swedish Research Council (Vetenskapsr\aa{}det)
  under Grant
  No.~621-2014-4864 \emph{Termination Certificates for Dependently-Typed Programs and Proofs via Refinement Types}.
  The first author is grateful for recent discussions with Thierry Coquand, Nils Anders Danielsson, and Sandro Stucki
  which helped clarifying the thoughts leading to this work.
  He also acknowledges past discussions with Christoph-Simon Senjak.
  The incentive to write this article came during the EU Cost Action CA15123 EUTYPES meeting in Ljubljana in January 2017; thanks to Andrej Bauer for organizing it.
\end{acks}
